Bayesian non-parametric inference for multivariate peaks-over-threshold models

### 2023-06-14 methodology_projgamma.tex

The projected gamma family is simple to specify and has very tractable 
  computational properties. Thus, we use it as a building block for the 
  angular measure $\Phi$ models. To build a flexible family of distributions 
  in ${\mathbb S}_p^{d-1}$ we consider mixtures of projected gamma densities 
  defined as
  \begin{equation} \label{eqn:PGmix}     
     f(\bm{y}) = \int_\Theta \mathcal{PG}(\bm{y}\mid \bm{\theta}) dG(\bm{\theta}), 
  \end{equation} 
where $\bm{\theta} = (\bm{\alpha}, \bm{\beta})$. Following a Bayesian 
  non-parametric approach \citep{Ferguson74,Antoniak1974} we assume that $G$ 
  is drawn from a random measure. In particular, assuming a Dirichlet process 
  prior for $G$, we have a hierarchical formulation of the mixture model that, 
  for a vector of observations $\bm{y}_i$, is given by
  \begin{equation}\label{eqn:dppg}
    \begin{aligned}
    \bm{y}_i &\sim \text{PG}(\bm{y}_i\mid \bm{\theta}_i)\\
    \bm{\theta}_i &\sim G
    \end{aligned}
    ~\hspace{1cm}
    \begin{aligned}
    G &\sim \mathcal{DP}(\eta, G_0),
    \end{aligned}
  \end{equation}
  where $\mathcal{DP}$ denotes a Dirichlet process, $\eta$ is the precision 
  parameter, and $G_0$ is the centering distribution.

@@

    &\propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i\ell}\right)^{\alpha_{j\ell} - 1}\alpha_{j\ell}^{\xi_\ell - 1}e^{-\tau_\ell \alpha_{j\ell}}\times}{\Gamma^{n_j}(\alpha_{j\ell})} \\

@@

To speed convergence and effectively explore the parameter space, we employ parallel 
tempering \citep{earl2005} for the log-normal models. This technique runs parallel
chains at ascending temperatures--that is, for chain $s$, the posterior density is 
exponentiated to the reciprocal of temperature $t_s$.  For the \emph{cold} chain, 
$t_1 := 1$.  Let $E_s$ be the log-posterior density under the current parameter 
state for chain $s$.  Then states between chains $r,s$ are exchanged via a Metropolis 
step with probability 
\[
\alpha_{rs} = \min
\left[1, \exp\left\lbrace(t_{r}^{-1} - t_{s}^{-1})(E_r - E_s)\right\rbrace\right].
\]
Higher temperatures serve to \emph{flatten} the posterior distribution, increasing the 
speed at which the chain explores parameter space.

Convergence of the MCMC chain is assessed by stabilization of the log-posterior 
density. For all of our datasets, none of the models required more than 40 thousand 
iterations to achieve stability, so we use that as a burn-in, and thereafter sample 
an additional 10 thousand iterations. We then thin the chain by 10 to collect 1000 
total samples with which to generate posterior predictive densities.

>>>>

The projected gamma family is simple to specify and has very tractable 
  computational properties. Thus, we use it as a building block for the 
  angular measure $\Phi$ models. To build a flexible family of distributions 
  in ${\mathbb S}_p^{d-1}$ we consider mixtures of projected gamma densities 
  defined as
  \begin{equation} \label{eqn:PGmix}     
     f(\bm{y}) = \int_\Theta \mathcal{PG}(\bm{y}\mid \bm{\theta}) dG(\bm{\theta}), 
  \end{equation} 
where $\bm{\theta} = (\bm{\alpha}, \bm{\beta})$. Following a Bayesian 
  non-parametric approach \citep{Ferguson74,Antoniak1974,muller2015},
  % \st{\bruno{\bf Peter:
  % Please provide a more recent and  approacheable reference, possibly a book. If you can't
  % find one  ask Thanos or Juhee, our BNP experts.} 
  we assume that $G$ 
  is drawn from a random measure. In particular, assuming a Dirichlet process 
  prior for $G$, we have a hierarchical formulation of the mixture model that, 
  for a vector of observations $\bm{y}_i$, is given by
  \begin{equation}\label{eqn:dppg}
    \begin{aligned}
    \bm{y}_i &\sim \text{PG}(\bm{y}_i\mid \bm{\theta}_i)\\
    \bm{\theta}_i &\sim G
    \end{aligned}
    ~\hspace{1cm}
    \begin{aligned}
    G &\sim \mathcal{DP}(\eta, G_0),
    \end{aligned}
  \end{equation}
  where $\mathcal{DP}$ denotes a Dirichlet process, $\eta$ is the precision 
  parameter, and $G_0$ is the centering distribution.

@@

    &\propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i\ell}\right)^{\alpha_{j\ell} - 1}\alpha_{j\ell}^{\xi_\ell - 1}e^{-\tau_\ell \alpha_{j\ell}}}{\Gamma^{n_j}(\alpha_{j\ell})} \\

@@

To effectively employ the sample space for a joint Metropolis step, as well as
    speeding convergence, we employ a parallel tempering algorithm\citep{earl2005} 
    for the log-normal models. This technique runs parallel MCMC
    chains at ascending temperatures--that is, for chain $s$, the posterior density is 
    exponentiated to the reciprocal of temperature $t_s$.  For the \emph{cold} chain, 
    $t_1 := 1$.  Let $E_s$ be the log-posterior density under the current parameter 
    state for chain $s$.  Then states between chains $r,s$ are exchanged via a Metropolis 
    step with probability 
    \[
    \alpha_{rs} = \min
    \left[1, \exp\left\lbrace(t_{r}^{-1} - t_{s}^{-1})(E_r - E_s)\right\rbrace\right].
    \]
    Higher temperatures serve to \emph{flatten} the posterior distribution, meaning hotter
    chains have a higher probability of making a given transition, or will make larger 
    transitions.  As such, they will more quickly explore the parameter space, and share
    information gained through state exchange.

### 2023-06-14 results.tex

\section{Data illustrations\label{sec:results}} 
\makenote{this title seems weird?}

@@

Figure~\ref{fig:simpples} shows the average rise over baseline in energy score as
    calculated on $\mathbb{S}_{\infty}^{d-1}$ using the kernel metric described
    in Proposition~\ref{prop:rot}, for models trained on simulated data.  After training
    a model, a posterior predictive dataset is generated, and the energy score is
    calculated as a Monte Carlo approximation of Equation~\ref{eq:es}.
    The \emph{baseline} value is the energy score of a new dataset from the same
    generating distribution as the training dataset, evaluated against the training 
    dataset.  We considered a total of four variations of the Dirichlet-process mixture
    of projected gammas model: whether to restrict $\beta_{j\ell} := 1$ as the projected
    \emph{restricted} gamma (DP-PRG) akin to a projected Dirichlet, or allow them to be 
    learned with a gamma prior (DP-PG); and for the shape parameters whether to use a 
    gamma centering distribution (-G) or a log-normal centering distribution (-LN).
    For comparison we also included two other distributions on $\mathbb{S}_{1}^{d-1}$:
    The Dirichlet distribution (with gamma priors), and the pairwise betas distribution
    developed in~\cite{COOLEY2010}.  Note that the pairwise betas model respects
    the moment restriction generally assumed.\makenote{This phrasing needs to be changed.
    The point I am trying to make is that because the moment-restriction assumption is 
    discarded, a moment-restricted model is inadequate, as is borne out in the
    simulation study.}  We observe that the projected gamma models dominate, and within
    that class of model, the choice of centering distribution is irrelevant.  For the
    projected \emph{restricted} gamma models, the choice of a multivariate log-normal 
    centering distribution negatively affects model fidelity.

@@

Table~\ref{tab:dev} shows the values of the estimated posterior predictive loss and energy 
    scores for the different models considered. We observe that, as in the case of the simulation 
    study shown in Figures~\ref{fig:simpples}, the preferred model is the projected restricted 
    gamma models, though for the lower-dimensional ERA-Interim data, all models perform comparably.
    Interestingly, this preference for the restricted gamma form is the opposite of the result 
    obtained in the simulation example.

>>>>

\section{Data illustrations\label{sec:results}} 
We apply the aforementioned models to generated angular data, as well as a data analysis
    of atmospheric data.
    Assessing convergence of an MCMC chain for a large-dimensional model is difficult.
    For our part, we assess convergence by stabilization of the log-posterior 
    density. For all of our datasets, none of the models required more than 40 thousand 
    iterations to achieve stability, so we use that as a burn-in, and thereafter sample 
    an additional 10 thousand iterations. We then thin the chain by 10 to collect 1000 
    total samples with which to generate posterior predictive densities.
    Implementation details play a role here: posterior samples from the gamma-centered 
    models are moderately autocorrelated as compared to the log-normal centered models.
    This is a benefit of the parallel tempering algorithm employed in fitting the 
    log-normal models, but parallel tempering necessitates an additional computational cost,
    growing linearly with the number of temperatures.  We used a temperature ladder
    of $t_s = 1.3^{s}$, for $s \in \lbrace0,1,\ldots,5$.  This was verified to produce
    acceptable swap probabilities both for the simulated data, and real data.

Our hyperprior parameters are set as follows: for the gamma-centered models (PG--G, PRG--G), 
    the shape parameter for the centering 
    distribution~$\xi_{\ell}\sim \mathcal{G}\left(1,1\right)$, and
    rate parameter~$\tau_{\ell}\sim\mathcal{G}\left(2,2\right)$.
    For the log-normal centered models (PG--LN, PRG--LN), the centering distribution's
    log-mean $\bm{\mu}\sim\mathcal{N}_d\left(0,\bm{I}_d\right)$, and covariance 
    matrix~$\Sigma\sim\mathcal{IW}\left(d + 10, (d+10)\bm{I}_{d}\right)$.  These values are 
    intended such that draws from the prior for $\Sigma$ will weakly tend towards the 
    identity matrix.
    For models learning rate parameters $\beta_{j\ell}$ (PG--G, PG--LN), the centering
    distribution's shape parameter $\zeta\sim\mathcal{G}\left(1,1\right)$ and rate parameter
    $\sigma\sim\mathcal{G}\left(2,2\right)$.  The choice of $\mathcal{G}(2,2)$ for rate
    parameters was meant to draw estimates for the value away from 0 for numerical stability.

@@

Figure~\ref{fig:simpples} shows the average rise over baseline in energy score as
    calculated on $\mathbb{S}_{\infty}^{d-1}$ using the kernel metric described
    in Proposition~\ref{prop:rot}, for models trained on simulated data.  After training
    a model, a posterior predictive dataset is generated, and the energy score is
    calculated as a Monte Carlo approximation of Equation~\ref{eq:es}.
    The \emph{baseline} value is the energy score of a new dataset from the same
    generating distribution as the training dataset, evaluated against the training 
    dataset.  We considered a total of four variations of the Dirichlet-process mixture
    of projected gammas model: whether to restrict $\beta_{j\ell} := 1$ as the projected
    \emph{restricted} gamma (DP-PRG) akin to a projected Dirichlet, or allow them to be 
    learned with a gamma prior (DP-PG); and for the shape parameters whether to use a 
    gamma centering distribution (-G) or a log-normal centering distribution (-LN).
    For comparison we also included two other distributions on $\mathbb{S}_{1}^{d-1}$:
    The Dirichlet distribution (with gamma priors), and the pairwise betas distribution
    developed in~\cite{COOLEY2010}.  
    % Note that the pairwise betas model respects
    % the moment restriction generally assumed.\makenote{This phrasing needs to be changed.
    % The point I am trying to make is that because the moment-restriction assumption is 
    % discarded, a moment-restricted model is inadequate, as is borne out in the
    % simulation study.} \bruno{\bf I would avoid getting into the details of what the
    % pairwise beta model does. More important is to say why we chose it, and how available 
    % it is} 
    We observe that for simulated data using Algorithm~\ref{algo:simulated}, the 
    projected gamma models dominate, and within that class of model, the choice of 
    centering distribution is irrelevant.  For the projected \emph{restricted} 
    gamma models, the choice of a multivariate log-normal centering distribution 
    appears to negatively affect model fidelity.

@@

% Table~\ref{tab:dev} shows the values of the estimated posterior predictive loss and energy 
%     scores for the different models considered. We observe that, as in the case of the simulation 
%     study shown in Figures~\ref{fig:simpples}, the preferred model is the projected restricted 
%     gamma models, though for the lower-dimensional ERA-Interim data, all models perform comparably.
%     Interestingly, this preference for the restricted gamma form is the opposite of the result 
%     obtained in the simulation example.
Table~\ref{tab:dev} shows the values of the estimated posterior predictive loss and energy 
    scores for the different models considered. We observe that, contrary to the results in the
    simulation study in Figure~\ref{fig:simpples}, the preferred model is the projected restricted 
    gamma models, though for the lower-dimensional ERA-Interim data, all models perform comparably.
    Interestingly, this preference for the restricted gamma form is the opposite of the result 
    obtained in the simulation example.
